The pollination trade-off

### 2018-03-01 publication/manuscript.Rmd
Initial introduction

@@

```{r fig-model-results-global, fig.height = 2, fig.cap = "Distribution of effect estimates for models of conspecific and heterospecific pollen density gain"}
loadd(tidied_fixed, sites)

tidied_fixed %>%
  filter(term != "(Intercept)" , 
         var_trans == 'log',
         scale == "global"
         ) %>%
  group_by(pollen_category,scale, model, var_trans, term) %>%
  summarise(estimate = first(estimate)) %>%
  group_by() %>%
  humanize() %>%
  ggplot(aes(x = estimate, 
             colour = pollen_category, 
             fill = pollen_category)) +
  geom_vline(xintercept = 0, linetype = 2, size = 0.25) +
  stat_density(geom = 'area',
               alpha = 0.15,
               position = position_identity(),
               draw_quantiles = c(0.5, 0.95, 0.05), 
               trim = F, 
               scale = "count",
               bins = 15) +
  pub_theme() +
  facet_grid( ~ term, scales = "free_x", space = "free_x") +
  scale_color_manual(values = colour_scale_values) +
  scale_fill_manual(values = colour_scale_values) +
  scale_x_continuous(breaks = seq(-2,2, by = 0.25)) + 
  theme(legend.position = "top") +
  labs(colour = "", fill = "", x = "estimated effect on pollen gain")
```

```{r fig-trade-off-predictions, fig.width=3.5, fig.height= 2.5, fig.cap = "Trade off predictions"}
loadd(trade_off_predictions)

trade_off_predictions %>%
  mutate_at(c("conspecific", "heterospecific"), exp) %>%
  mutate(gain = conspecific / (conspecific + heterospecific)) %>%
  filter(!(value > 2.2 & term == "k")) %>%
  humanize() %>%
  ggplot(aes(x = value, y = gain, colour = term)) +
  geom_hline(yintercept = 0.5, size = 0.25, linetype = 2) + 
  geom_point(shape = 21, alpha = 0.5, size = 1) +
  geom_smooth(aes(fill = term), 
              se = F, size = 0.5, method = "glm", 
              method.args = list(family = "binomial"), alpha = 0.1)+
  pub_theme() +
  scale_color_manual(values = colour_scale_values) +  
  scale_fill_manual(values = colour_scale_values) +
  theme(legend.position = "top") +
  labs(colour = "", fill = "", 
       y = "scaled trade-off\nconspecific - heterospecific pollen gain", 
       x = "scaled variable value")
```

>>>>

Pollination systems play a disproportionally important role in food production and maintenance of global biodiversity. 
In the context of animal pollination, the support to biodiversity is fundamentally achieved by virtue of intra-species facilition—the positive feedback loops that exist between plants that share pollinators or pollinators that share plants. 
Facilitation is able to promote species coexistence because it offsets the effects of direct competition for resources. 
Indeed, studies that are fundamental to our current knowledge of mutualism predict that an upper limit to biodiversity is achieved when the number of mutualistic partners is maximised in a community. 
And yet, fully connected natural pollination communities do not exist.

This is so, because sharing a mutualistic partner also has costs. 
The end result is that, when the cummulative costs have negative impact on fitnessis, species also start competeing for mutualistic partners. 
For the 88% of flowering plants that rely on animal pollination, there are two mechanisms trough which competition for pollinators might arrise. 
The first is competition for visits, where plants with more atractive flowers reduce the pollination service for those less atractive. 
The second is interspecific pollen transfer, where even receiving a visit might not translate into net facilitation. 
A focal plant might receive pollen from a different species, or conversely pollen from the focal plan might be lost to different species. 
Here we focus on this second mechanism.

The potential negative effects of competition for pollinators have been long recognised and widely documented. 
It is currently clear that they have the potential to drive the evolution of flower phenologies, morphologies and reproductive strategies as well as driving ecological diferentiation. 
Indeed there is considerable evidence demonstrating the effects that interspecific pollen transfer in two-species systems. 
However, these competitive interactions between species pairs do not occur in isolation. 
Instead, plants often ocurr in communities in which multiple competitive interactions operate simultaneusly and lead to emergent phenomena not observed at smaller scales.

Nevertheless, only a handful of studies have examined interspecific pollen transfer at the community level, and therefore we still understand little about how compeition for pollinators operates at this scale. 
All together, the few existing studies show that, as expected, the effective result of mutualistic interactions is, in most cases, one of facilitation. 
Importantly, they also show a large variability in this result—both across species and across communities. 
However, the driving factors behind this variability are currently not understood.

But understanding what makes species and communities different is important not only because the ecological and evolutionary implications. 
It is also important because it is fundamentaly related to the way we understand and mathematically represent mutualistic systems. 
An important proportion of widely used theoretical models of pollination account for the negative effects of mutualisim by shaping the benefits

It is generally believed that there is a trade-off between the number of mutualistic parters a plant species has, and the strength of its interactions (mutualistic benefit).
The trade-off arises from the premise that an increase on the number of mutualistic partners might increase facilitation, but might also increase competition for pollinators because it can lead to a reduction of the proportion of conspecific pollen deposited on the stigmas.
Species properties and the environmental context seem to influence whether sharing pollinators leads to facilitation or competition [@Tur2016a], however the specific mechanisms behind these observations are poorly understood.
Furthermore, recent theoretical advances that explicitly include the mutualistic trade-off in pollination models have highlighted its importance.
Indeed, it has been shown that the trade-off is related to the feasibility and stability of mutualistic communities [@Rohr2014], the species' response to change [@Saavedra2013], the onset of critical transitions [@Dakos2014a], and the spread of perturbations [@Suweis2015].
However, The extent to which the trade-off operates in nature and its relationship with facilitation-competition dynamics has not been empirically evaluated.
Here, we used data on the quality and quantity of pollination in six pollination communities to quantify the empirical support for the mutualistic trade-off concept and identify the mechanisms that structure it.
Contrasting to our initial expectation, we found that the mutualistic benefit is only weakly related to the number of pollination partners.
We found that this relationship is weak because plants with a large number of shared partners are able to reduce the cost of sharing by increasing the niche partitioning.
Importantly, the niche partitioning strategies are modulated by plant abundance and are not homogeneously employed by all plant species.

@@

```{r fig-model-results-global, fig.height = 2, fig.cap = "Distribution of effect estimates for models of conspecific and heterospecific pollen density gain. Results are shown for models with explanatory var"}
loadd(tidied_fixed, sites)

tidied_fixed %>%
  filter(term != "(Intercept)" , 
         var_trans == 'log',
         scale == "global"
         ) %>%
  group_by(pollen_category,scale, model, var_trans, term) %>%
  summarise(estimate = first(estimate)) %>%
  group_by() %>%
  humanize() %>%
  ggplot(aes(x = estimate, 
             colour = pollen_category, 
             fill = pollen_category)) +
  geom_vline(xintercept = 0, linetype = 2, size = 0.25) +
  stat_density(geom = 'line',
               # alpha = 0.15,
               position = position_identity(),
               draw_quantiles = c(0.5, 0.95, 0.05), 
               trim = T, 
               scale = "count",
               bins = 15) +
  pub_theme() +
  facet_grid( ~ term, scales = "free_x", space = "free_x") +
  scale_color_manual(values = colour_scale_values) +
  scale_fill_manual(values = colour_scale_values) +
  scale_x_continuous(breaks = seq(-2,2, by = 0.25)) + 
  theme(legend.position = "top") +
  labs(colour = "", fill = "", x = "estimated effect on pollen gain")
```

The gain in conspecific pollen is positively correlated with the gain in heterospecific pollen. 
The slope of the relationship is such that it is more likely that the gain on heterospecific pollen increases faster than the gain on conspecific pollen (Figure \@ref(fig:fig-trade-off-predictions)A). 
In addition, we found that from all the explanatory variables examined, only the plant's relative abbundance has a positive effect on the proportion of conspecific pollen gained (Figure \@ref(fig:fig-trade-off-predictions)B). 
For most of the parameter space plants obtain a larger proportion of heteropspecific than conspecific pollen. 
However, plants can obtain a larger proportion of conspecific pollen when they have high abundance and to a smaller extent when they have a low degree.

```{r fig-trade-off-predictions, fig.height= 2.5, fig.cap = "Trade off predictions j."}
loadd(trade_off_predictions, model_linear_fits, tidied_fixed)
# loadd(model_correlations)

# p1 <- model_correlations %>%
#   filter(var_trans == "log", 
#          scale == "global") %>%
#   ggplot(aes(x = pearson_slope)) + 
#   geom_vline(xintercept = 0.5, size = 0.25, linetype = 2) +
#   stat_density(trim = F, geom = "line") +
#   pub_theme() +
#   labs(x = "slope\n(heterospecific ~ conspecific pollen gain)")
# 
get_pred_range <- function(x){
  tidied_fixed %>% filter(pollen_category == x, var_trans == "log", scale == "global") %$% estimate %>% range()
}

ab_lines <- model_linear_fits %>%
  filter(var_trans == "log",
         scale == "global") %>%
  spread(sma_parameter, value)

p1 <- data_frame(x = get_pred_range("heterospecific"),
                 y = get_pred_range("conspecific")) %>%
  ggplot(aes(x, y)) +
  geom_abline(data = ab_lines, 
              aes(slope = slope, intercept = elevation), 
              size = 0.1, linetype = 1, colour = "grey", 
              alpha = 0.5) +
  geom_abline(data = summarise_if(ab_lines, is.numeric, median), 
              aes(slope = slope, intercept = elevation), 
              size = 0.5, linetype = 1, colour = "black") +
  geom_abline(slope = 1, intercept = 0, size = 0.25, linetype = 2) +
  geom_point(alpha = 0) +
  coord_equal(xlim = get_pred_range("heterospecific"),
                  ylim = get_pred_range("conspecific")) +
  pub_theme() +
  labs(x = "gain in heterospecific pollen", y = "gain in conspecific pollen")

p2 <- trade_off_predictions %>%
  mutate_at(c("conspecific", "heterospecific"), exp) %>%
  mutate(gain = conspecific / (conspecific + heterospecific)) %>%
  filter(!(value > 2.2 & term == "k")) %>%
  humanize() %>%
  ggplot(aes(x = value, y = gain, colour = term)) +
  geom_hline(yintercept = 0.5, size = 0.25, linetype = 2) + 
  geom_point(shape = 21, alpha = 0.5, size = 1) +
  geom_smooth(aes(fill = term), 
              se = F, size = 0.5, method = "glm", 
              method.args = list(family = "binomial"), alpha = 0.1)+
  pub_theme() +
  scale_color_manual(values = colour_scale_values) +  
  scale_fill_manual(values = colour_scale_values) +
  scale_y_continuous(labels = scales::percent) +
  theme(legend.position = "top") +
  labs(colour = "", fill = "", 
       y = "proportion conspecific pollen gain", 
       x = "scaled variable value")

cowplot::plot_grid(p1, p2, align = "h", axis = "tb", ncol = 2, labels = LETTERS[1:2], label_size = 8, rel_widths = c(0.35, 0.65))
```
